Galois Slicing as Differentiation

### 2025-07-11 main/introduction.tex
[ tex ] squash lattices
Tweak intro.

Existing approaches to program slicing are often tied to particular programming languages or implementations.
In this paper we develop a general categorical approach to program slicing, focusing on a particular technique
called \GPS, where the set of slices of a given value form a lattice of approximations and the forward and
backward slicing procedures generate Galois connections between these lattices. Our main contribution is that this
approach can be seen as a generalised form of differentiable programming, with slices of elements playing the
role of tangents. Our categorical approach should provide a suitable setting for exploring data provenance for
a variety of programming languages, and is easily configured to use alternative approximation strategies,
including quantitative forms of slicing.

@@

  Now suppose that for a given run of the program, we are interested in which of the numerical parts of the input are used to compute the output for the query parameters $l = \mathsf{a}$ and $l = \mathsf{b}$. We can use \GPS to do this. We arrange for the approximations of the input to form the following lattice, where the actual piece of data is at the top and information lost by approximation is represented by $\bot$s:
  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      % row 2
      \node [below of=top] (ioi) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [left of=ioi,xshift=-3cm] (oii) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      \node [right of=ioi,xshift=3cm] (iio) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      % row 3
      \node [below of=ioi] (oio) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      \node [left of=oio,xshift=-3cm] (ooi) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [right of=oio,xshift=3cm] (ioo) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};
      % row 4
      \node [below of=oio] (bot) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
      \draw (oii) -- (ooi);
      \draw (oii) -- (oio);
      \draw (iio) -- (ioo);
      \draw (iio) -- (oio);
      \draw (ooi) -- (bot);
      \draw (oio) -- (bot);
      \draw (ioo) -- (bot);
    \end{tikzpicture}
  \end{center}
  In both runs of the program, the output approximation lattice looks like this, where $1$ is the actual data point that was returned, and $\bot$ indicates that we are approximating this piece of data away:
  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$1$};
      \node [below of=top] (bot) {$\bot$};
      \draw (top) -- (bot);
    \end{tikzpicture}
  \end{center}
  These are not the only choices of approximation lattices that we could have made. For the input, we have chosen a lattice that allows us to ``forget'' (approximate away) numbers in the input, but not any of the other data or the overall structure. However, other choices are also useful. Indeed, one of the aims of this work is to clarify how to choose an approximation structure appropriate for different tasks by use of type information. We elaborate on this further in \secref{models-of-total-gps}.

>>>>

Existing approaches to program slicing are often tied to particular programming languages or implementations.
In this paper we develop a general categorical approach to program slicing, focusing on a particular technique
called \GPS, where the set of slices of a given value form a lattice of approximations and the forward and
backward slicing procedures generate Galois connections between these lattices. Our main contribution is that
this approach can be seen as a generalised form of automatic differentiation, with slices of values playing
the role of tangents. Our categorical approach should provide a suitable setting for enabling ``automatic''
data provenance for a variety of programming languages, and is easily configured to use alternative
approximation strategies, including quantitative forms of slicing.

@@

  Now suppose that for a given run of the program, we are interested in which of the numerical parts of the input are used to compute the output for the query parameters $l = \mathsf{a}$ and $l = \mathsf{b}$. We can use \GPS to do this. We arrange for the approximations of the input to form the following lattice, where the actual piece of data is at the top and information lost by approximation is represented by $\bot$s:
  \begin{center}
    \begin{tikzpicture}[node distance=0.75cm]
      \node (top) at (0,0) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      % row 2
      \node [below of=top] (ioi) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [left of=ioi,xshift=-3cm] (oii) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, 1)]$};
      \node [right of=ioi,xshift=3cm] (iio) {$[(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      % row 3
      \node [below of=ioi] (oio) {$[(\mathsf{a}, \bot), (\mathsf{b}, 1), (\mathsf{a}, \bot)]$};
      \node [left of=oio,xshift=-3cm] (ooi) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, 1)]$};
      \node [right of=oio,xshift=3cm] (ioo) {$[(\mathsf{a}, 0), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};
      % row 4
      \node [below of=oio] (bot) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
      \draw (oii) -- (ooi);
      \draw (oii) -- (oio);
      \draw (iio) -- (ioo);
      \draw (iio) -- (oio);
      \draw (ooi) -- (bot);
      \draw (oio) -- (bot);
      \draw (ioo) -- (bot);
    \end{tikzpicture}
  \end{center}
  In both runs of the program, the output approximation lattice looks like this, where $1$ is the actual data point that was returned, and $\bot$ indicates that we are approximating this piece of data away:
  \begin{center}
    \begin{tikzpicture}[node distance=0.75cm]
      \node (top) at (0,0) {$1$};
      \node [below of=top] (bot) {$\bot$};
      \draw (top) -- (bot);
    \end{tikzpicture}
  \end{center}
  These are not the only choices of approximation lattices that we could have made. For the input, we have chosen a lattice that allows us to ``forget'' (approximate away) numbers in the input, but not any of the other data or the overall structure. However, other choices are also useful. Indeed, one of the aims of this work is to clarify how to choose an approximation structure appropriate for different tasks by use of type information. We elaborate on this further in \secref{models-of-total-gps}.
